Gap formation and stability in non-isothermal protoplanetary discs

### 2015-03-05 paper/response.txt

We thank the reviewer for a timely report, and apologize for the delay
in our response. We have performed additional simulations to address
some of the reviewer's questions. In particular, Fig. 13 has been
expanded to include more cases to strengthen our main result 
of a non-monotonic dependence of the vortex lifetime on beta. The
revisions to the manuscript are made in bold, and further explained
below.

- §§4 and 5: would the results of a simulation with a very short
  cooling time (beta << 1) be the same as those of a simulation with a
  locally isothermal equation of state and same unperturbed sound
  speed? 
  ********************************************************************
  We have performed additional (lower resolution) simulations that
  compare beta=0.01 with a locally isothermal disks, which we obtain 
  from our numerical models by setting gamma=1.01. We find a slightly
  smaller linear growth rate in the locally isothermal case. However,
  the vortex formation and evolution are similar.

  We obtain longer gap-edge vortex lifetimes with beta=0.01 than 
  beta=0.1. We now include this result in Fig. 13 (all simulations in
  this figure employ the same resolution).

@@

  ********************************************************************

- §4.4: why does the vortex’s perturbed density decrease on the long
  term? I suppose a related question is: what controls the long-term
  evolution of the vortex’s potential vorticity or entropy-modified
  potential vorticity? (whichever quantity is relevant). 
  ********************************************************************
  In theory a vortex should live forever in 2D inviscid flow. In
  practice, however, there is numerical viscosity which we suspect is
  responsible for the long term decay. As the vortex weakens
  (elongates) its area increases and its surface density drops. In
  addition, for beta=0.1, 0.5, 1.0, for which the initially-formed
  vortex is stronger, we observe the appearance of spiral waves
  associated with the vortex, which may contribute to the decay. It
  can be seen in Fig. 4 that these cases experience a faster decay
  rate.

@@

  ********************************************************************

- §5.2-5.3: here again I’d like to have more details on the vortex
        evolution and decay: 	
        . Do the authors see the vortex migrate? I did a simulation
        with the authors’ setup and a rather long cooling time
        (beta_tilde ~ 8) where I see the vortex moving from ~6.5 rh to
        ~5.3 rh relative to the planet’s orbital radius before the
        vortex starts decaying (rh = Hill radius). The time when the
        vortex starts decaying actually corresponds to a rather sudden
        increase in the planet-vortex distance, which goes up to ~6.8
        rh until the vortex dies off completely. If the authors get a
        similar migration behaviour in their simulations, maybe this
        could give further insight into the vortex evolution (vortex
        getting stronger during the quasi steady state before fading
        away). The connection between migration and the appearance of
        spiral waves launched by the vortex could also be made. 
	********************************************************************
        We thank the reviewer for raising the issue of vortex
        migration. We have re-examined our simulation results. We
        indeed observe vortex migration but our result is somewhat
        different to that of the reviewer.

 	We do not see significant vortex migration during quasi-steady
 	state. This may be due to the fact that the 	  
 	vortex forms at a surface density maximum, which is able to
 	trap vortices (Paardekooper et al 2010). Thus, in our
 	simulations the growth during quasi steady state is not due to
 	vortex migration.

	Instead, we observe the vortex begins to migrate inwards
	with the appearance of the spiral wakes (cf. planetary
	migration). However, we find the vortex eventually migrates
	outward to about 6.5rh before complete decay.

@@

	We remark that for a vortex formed at a surface density
 	maximum (e.g. from the RWI), that subsequent perturbations may
 	lead to inwards or outwards vortex migration, depending on
 	what side of the maximum the vortex is displaced to
 	(Paardekooper et al 2010).

	Although we do not understand the discrepancy between our
	results and the reviewer's, we do agree with the reviewer
	that vortex migration probably plays a role in determining
	the overall vortex lifetime (it may affect the duration of the
	quasi-steady state and/or the decay timescale). We now mention
	this in the discussions and caveats.

@@

        ********************************************************************

        . During the quasi steady state the vortex grows as a result
        of continuous vortensity generation upon crossing the planet’s
        wake. What about possible turbulent diffusion of potential
        vorticity if the planet forms in a region of the disc with
        non-negligible ‘turbulent viscosity’? Have the authors done
        simulations with a non vanishingly small viscosity? 
        ********************************************************************
	Since a low-viscosity disc is needed for the RWI to operate,
        we have chosen to focus on inviscid simulations. This also
        helps to minimize complications from vortex decay directly due
        to a physical  
        viscosity (which also contributes to heating).

### 2015-03-05 paper/model.tex
revised manuscript, submitted

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum equations, $\bm{f}_\nu$ represent viscous forces, {\bf
which includes artificial viscosity to handle shocks, and a
Navier-Stokes viscosity} whose magnitude is  
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$.

\subsection{Heating and cooling}
In the energy equation, the heating term $\mathcal{H}$ is defined as 
\begin{align}
  \mathcal{H} \equiv Q^+ - Q^+_i\frac{\Sigma}{\Sigma_i}, 
\end{align}
where $Q^+$ represents viscous heating {\bf(from both physical and
  artificial viscosity)} and subscript $i$ denotes
evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
\begin{align}
  \mathcal{C} \equiv \frac{1}{t_c}\left(e -
  e_i\frac{\Sigma}{\Sigma_i}\right),  
\end{align}
where $t_c = \beta\Omega_k^{-1}$ is the cooling time,
$\Omega_k=\sqrt{GM/r^3}$ is the Keplerian frequency and $\beta$ is an
input parameter. This cooling prescription allows one 
to explore the full range of thermodynamic response of the disc in a 
systematic way: $\beta\ll1$ is a locally isothermal disc while
$\beta\gg1$ is an adiabatic disk.

>>>>

The total potential $\Phi$ includes the stellar potential, planet potential
(described below) 
and indirect potentials to account for the non-inertial reference
frame. 
In the momentum equations, $\bm{f}_\nu$ represent viscous forces, {\bf
which includes artificial bulk viscosity to handle shocks, and a
Navier-Stokes viscosity} whose magnitude is  
characterised by a constant kinematic viscosity parameter
$\nu$. However, we will be considering effectively inviscid discs by
adopting small values of $\nu$.

\subsection{Heating and cooling}
In the energy equation, the heating term $\mathcal{H}$ is defined as 
\begin{align}
  \mathcal{H} \equiv Q^+ - Q^+_i\frac{\Sigma}{\Sigma_i}, 
\end{align}
where $Q^+$ represents viscous heating {\bf(from both physical and
  artificial viscosity)} and subscript $i$ denotes
evaluation at $t=0$. The cooling term $\mathcal{C}$ is defined as
\begin{align}
  \mathcal{C} \equiv \frac{1}{t_c}\left(e -
  e_i\frac{\Sigma}{\Sigma_i}\right),  
\end{align}
where $t_c = \beta\Omega_k^{-1}$ is the cooling time,
$\Omega_k=\sqrt{GM/r^3}$ is the Keplerian frequency and $\beta$ is an
input parameter. This cooling prescription allows one 
to explore the full range of thermodynamic response of the disc in a 
systematic way: $\beta\ll1$ is a locally isothermal disc while
$\beta\gg1$ is an adiabatic disc.

### 2015-03-05 paper/paper.tex
revision in the works
revised manuscript, submitted

\maketitle
\begin{abstract}
  Several observations of transition discs show lopsided
  dust-distributions. A potential explanation is the formation of a
  large-scale vortex acting as a dust-trap at the edge of a gap opened
  by a giant planet. Numerical models of gap-edge vortices have
  so far employed locally isothermal discs, but the 
  theory of this vortex-forming or `Rossby wave' instability was
  originally developed for adiabatic discs.  
  We generalise the study of planetary gap stability to non-isothermal
  discs using customised numerical simulations of disc-planet
  systems where the planet opens an unstable gap. 
  We include in the energy equation a simple cooling function with
  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
  the Keplerian frequency, and examine the effect of $\beta$ on the
  stability of gap edges and vortex lifetimes. We find increasing
  $\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
  dominant azimuthal wavenumber $m$ decreases.  
  We find a quasi-steady state consisting of one  
  large-scale, over-dense vortex circulating the outer gap edge, typically  
  lasting $O(10^3)$ orbits. 
  Vortex lifetimes were found to generally increase with cooling times up to an
  optimal value, beyond which vortex lifetimes decrease. This
  non-monotonic dependence is qualitatively consistent with 
  recent studies using strictly isothermal discs that vary the disc aspect ratio. 
%  presented in Fu et al. (2014, ApJL, 788, L41). 
\end{abstract}

@@

However, studies of Rossby vortices at planetary gap-edges have 
adopted locally isothermal discs, %cite lin11, zhu14, lyra
where the disc temperature is a fixed function of
position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. 
%isothermal valid for outer disc  
On the other hand, the theory of the RWI was in fact
developed for adiabatic discs \citep{li00}, which permits
heating. %cite 
In adiabatic discs, the relevant quantity for stability
becomes a generalisation of the PV that accounts for entropy variations
\citep{lovelace99}.

@@

For example, previous
linear simulations of the RWI found  
that increasing the sound-speed favours instability \citep{li00,lin13}.  
In the context of
planetary gaps, however, the increased temperature may also act to
stabilise the disc by making gap-opening more difficult. It is 
therefore of theoretical interest to clarify the effect of heating and
cooling on the stability of planetary gaps.

@@

This paper is organised as follows. In \S\ref{model} we describe the
equations governing the disc-planet system and initial conditions. Our
numerical approach, including diagnostic measures, are given in
\S\ref{method}. We present results from two sets of numerical
experiments. In \S\ref{linear1} we use disc-planet interaction to set
up discs with gaps, but study their stability without further
influence from the planet. %stability through background state 
We then perform long-term disc-planet simulations to examine the
lifetime of gap-edge vortices in \S\ref{nonlinear},  
as a function of the imposed cooling rate. We conclude and summarise
in \S\ref{summary} with a discussion of important caveats.

@@

\section*{Acknowledgments}
This project was initiated at the Canadian Institute for Theoretical
Astrophysics (CITA) 2014 summer student programme.  
Computations were performed on the GPC supercomputer at the
SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for
Innovation under the auspices of Compute Canada; the Government of
Ontario; Ontario Research Fund - Research Excellence; and the
University of Toronto.

>>>>

\maketitle
\begin{abstract}
  Several observations of transition discs show lopsided
  dust-distributions. A potential explanation is the formation of a
  large-scale vortex acting as a dust-trap at the edge of a gap opened
  by a giant planet. Numerical models of gap-edge vortices have
  so far employed locally isothermal discs, but the 
  theory of this vortex-forming or `Rossby wave' instability was
  originally developed for adiabatic discs.  
  We generalize the study of planetary gap stability to non-isothermal
  discs using customized numerical simulations of disc-planet
  systems where the planet opens an unstable gap. 
  We include in the energy equation a simple cooling function with
  cooling timescale $t_c=\beta\Omega_k^{-1}$, where $\Omega_k$ is
  the Keplerian frequency, and examine the effect of $\beta$ on the
  stability of gap edges and vortex lifetimes. We find increasing
  $\beta$ lowers the growth rate of non-axisymmetric perturbations, and the
  dominant azimuthal wavenumber $m$ decreases.  
  We find a quasi-steady state consisting of one  
  large-scale, over-dense vortex circulating the outer gap edge, typically  
  lasting $O(10^3)$ orbits. {\bf
    We find vortex lifetimes generally increase with the cooling
    timescale $t_c$} up to an optimal value {\bf of $t_c\sim 10$
    orbits}, beyond which vortex lifetimes decrease. This 
  non-monotonic dependence is qualitatively consistent with 
  recent studies using strictly isothermal discs that vary the disc
  aspect ratio.  {\bf The lifetime and observability of
    vortices in protoplanetary discs is therefore dependent on disc
    thermodynamics. }
%  presented in Fu et al. (2014, ApJL, 788, L41). 
\end{abstract}

@@

However, studies of Rossby vortices at planetary gap-edges have 
adopted locally isothermal discs, %cite lin11, zhu14, lyra
where the disc temperature is a fixed function of
position only \citep[e.g.][]{lyra08,lin11a,zhu14,fu14}. 
%isothermal valid for outer disc  
On the other hand, the theory of the RWI was in fact
developed for adiabatic discs \citep{li00}, which permits
heating. %cite 
In adiabatic discs, the relevant quantity for stability
becomes a generalization of the PV that accounts for entropy variations
\citep{lovelace99}.

@@

For example, previous
linear simulations of the RWI found  
that increasing the sound-speed favours instability \citep{li00,lin13}.  
In the context of
planetary gaps, however, the increased temperature may also act to
stabilize the disc by making gap-opening more difficult. It is 
therefore of theoretical interest to clarify the effect of heating and
cooling on the stability of planetary gaps.

@@

This paper is organised as follows. In \S\ref{model} we describe the
equations governing the disc-planet system and initial conditions. Our
numerical approach, including diagnostic measures, are given in
\S\ref{method}. We present results from two sets of numerical
experiments. In \S\ref{linear1} we use disc-planet interaction to set
up discs with gaps, but study their stability without further
influence from the planet. %stability through background state 
We then perform long-term disc-planet simulations to examine the
lifetime of gap-edge vortices in \S\ref{nonlinear},  
as a function of the imposed cooling rate. We conclude and summarize
in \S\ref{summary} with a discussion of important caveats.

@@

\section*{Acknowledgments}
This project was initiated at the Canadian Institute for Theoretical
Astrophysics (CITA) 2014 summer student programme. {\bf The authors
  thank an anonymous referee for an insightful report.} 
Computations were performed on the GPC supercomputer at the
SciNet HPC Consortium. SciNet is funded by: the Canada Foundation for
Innovation under the auspices of Compute Canada; the Government of
Ontario; Ontario Research Fund - Research Excellence; and the
University of Toronto.

### 2015-03-05 paper/results.tex
revised manuscript, submitted

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimenionlised by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

These `planet-off' simulations show that gap edges become more stable with
longer cooling times. This is expected because larger $\tilde{\beta}$
results in hotter gap profiles at $t=30P_0$ with less pronounced
generalised vortensity minima. Stabilisation with increased
cooling time is therefore due to a smoother basic state for the
instability, as it is more difficult for the planet to open a gap if
the disc is allowed to heat up.

@@

{\bf For completeness we also performed a lower resolution simulation
  where the sound-speed is effectively held fixed (a locally
  isothermal disk) by setting $\gamma=1.01$. This simulation
  yield a most unstable growth rate $\sigma/\Omega(r_p)\simeq 0.05$,
  compared with a value of $0.08$ for a case with
  $\tilde{\beta}=0.01$. However, the vortex evolutions were similar.}

@@

The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is comparable or shorter
than the instability growth time, $t_c\lesssim\gamma^{-1}$.  
Thus the disk cools back to its initial temperature corresponding to
$h=0.05$ before or during the instability growth, so we do not 
have a steady basic state to formulate a standard linear stability 
problem.

@@

{\bf This decay is probably due to numerical viscosity. During the
  slow decay the vortex elongates (weakens) while its radial width
  remains $O(H)$, so its surface density decreases. In addition, for 
  $\tilde{\beta}=0.1,\,0.5$ and $1.0$ we also observe the appearence of
  spiral waves associated with the vortex, which may contribute to
  its dissipation (see below).} 
% (true?) (For isothermal and long
% cases yes. Spiral waves appear in beta=0.1,0.5,1.0}) 
We will see in the
next section that this {\bf decay after linear growth} is very
different to when the planet potential is kept on.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1.1cm]{figures/longterm_planetoff}
  \caption{Long term simulations without the planet potential after
    the gap is set up. The $m=1$ surface density component,
    non-dimenionlised by the initial $m=0$ component, at the
    outer gap edge is shown as a function of the cooling timescale. 
  } \label{planetofflifetimeplot}
\end{figure}

@@

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time}\label{nonlinear} 
We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$. The planet potential is kept on
throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
for these 
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimise boundary
effects on vortex evolution, and apply open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$.

We comment that lower-resolution simulations with
$(N_r,N_{\phi})=(256,512)$ show similar behaviour and trends as the
high-resolution runs reported below. {\bf These simulations
  include a case with $\tilde{\beta}=0.01$ and a locally isothermal
  disk run, which show similar behavior. In particular, we find vortex
  lifetimes have a non-monotonic dependence on $\tilde{\beta}$, with
  longest lifetimes as $\tilde{\beta}\to 0$ (or locally isothermal
  disks) and for $\tilde{\beta}\sim 1$}.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/vortex_density}
  \caption{Average value of the relative surface density perturbation
    at vortex centres for various cooling
    times. Intial vortex overdensities decrease with cooling rate while
    vortex growth rates increase with cooling rate.
    \label{overdensity}}     
\end{figure}

@@

{\bf During the quasi-steady state the vortex 
  orbits at $r\sim r_p+6r_h$. We do not see significant
  vortex migration at this stage, since the vortex is located at a 
  surface density maximum \citep{paardekooper10}. However,
  simultaneous with the appearence of the wakes, we observe the vortex
  begins to migrate inwards to $r\sim r_p+5r_h$.  
  % This is a result of the the spiral waves transporting
  % angular momentum to the disk (Is this true? Shocks on the inner disk
  % seem stronger so I would think it would migrate outward). 
  }

@@

We also measured large increases in the Mach number near
the vortex as the $m=1$ surface density amplitude reaches maximum and begins to decay. 
Fig. \ref{machplot} plots the Mach number $M=|\bm{v} -
\bm{v}_\mathrm{vor}|/c_s$, where 
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre. 
During the quasi-steady state the Mach number increases 
steadily, and for all cases $M$ maximises about
% with larger growth rate for higher $\tilde\beta$.
$\sim 100P_0$ after the start the $m=1$ surface density
amplitude starts to decay. 
% {\bf check if  statement is true}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gapchange}
  \caption{Azimuthally averaged profiles of the relative surface
    density perturbations for  $\tilde\beta=1.0$
    before (solid) and after vortex decay
    (dashed). Alongside drastic reduction in the outer gap maxima, vortex decay
    smoothens out the sharp outer gap edge.
    \label{gap_smoothed}} 
\end{figure}

We characterise the smoothness of the outer gap edge with a dimensionless
gap edge gradient parameter
\begin{align}
  \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r)
    \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r)
    \rangle_{\phi}} \ra_{\Delta r}  
\end{align}
where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging, spanning
from centre of the gap to the radius of the surface density maximum.
A larger $\delta \Sigma$ characterises a sharper gap edge and  
larger local surface density maxima.

@@

This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet {\bf, and smooths out the outer surface density bump,    
} so the condition for the RWI 
becomes less favourable. {\bf In
  order to re-launch the RWI, the surface density bump should
  reform. However, this is difficult as there is no more material
  in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}) to
  form a surface density bump outside the gap.} {\bf This may} explain
why vortices do  not reform again (at least within the simulation timescale). 
{\bf After full decay the aspect ratio at the outer gap edge is
  $h\sim0.05$. }
% (0.06 for beta=5.0,10)}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gap_smoothness}
  \caption{Non-dimensional measure of the surface density gradient at
    the outer gap edge $\tilde\beta=1.0$. During the vortex quasi-steady state
    the gap edge is found to have a large gradient and sharp peak while vortex
    dissipation,
    which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
    works to smoothen out the gap edge.
% A large $\delta\Sigma$
 %   corresponds to a more sharply peaked outer gap edge.
    \label{smoothnessplot}}  
\end{figure}

@@

We now examine vortex lifetimes as function of the imposed cooling
times. {\bf For completeness additional simultions were done for $\tilde\beta=0.01,0.25,0.75,2.5,7.5$. }
 We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
at which the over-density of the vortex
% to reach $10\%$ of its max overdensity value
returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
%{\bf correct interpretation?}
 (which is on the order of the
initial over-density associated with the gap formation).

@@

For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
vortex lifetime is maximised for $\tilde{\beta}\to0$: we find 
$t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
$\tilde{\beta}=0.5$. This is due to the longer decay timescale with
decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
1$) the vortex lifetimes maximises at $\tilde\beta=2.5$ with 
$t_{\mathrm{life}}\approx 1650P_0$.  The other timescales $t_\mathrm{diss}$ and
$t_\mathrm{Mach}$ are less complicated, but still display a
non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
$\tilde{\beta}=1$. 
% his is in comparison with the timescale of intial vortex dissipation which has 
% non-monotonic dependence and a single maxima at $\tilde\beta=5.0$.
%non-monotonic dependence for all timescales

>>>>

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=1.2cm
  0cm 0cm 0cm]{figures/linear_stability}
  \caption{Evolution of azimuthal Fourier modes of disc surface
    density, non-dimensionalized by the initial axisymmetric component 
    $\Sigma_0(t=0)$, for the
    `planet-off' simulation with $\tilde{\beta}=10$. Colours correspond
    to different $m$ values. The $m=3$ component is the fastest growing
    mode during linear growth with a growth rate of 
    $\gamma=0.017\Omega(r_p)$.\label{linearmodes}}
  % {\bf is the $\Sigma_0$ at t=0?} }  
\end{figure}

@@

These `planet-off' simulations show that gap edges become more stable with
longer cooling times. This is expected because larger $\tilde{\beta}$
results in hotter gap profiles at $t=30P_0$ with less pronounced
generalised vortensity minima. Stabilization with increased
cooling time is therefore due to a smoother basic state for the
instability, as it is more difficult for the planet to open a gap if
the disc is allowed to heat up.

@@

{\bf For completeness we also performed a lower resolution simulation
  where the sound-speed is effectively held fixed (a locally
  isothermal disc) by setting $\gamma=1.01$. This simulation
  yield a most unstable growth rate $\sigma/\Omega(r_p)\simeq 0.05$,
  compared with a value of $0.08$ for a case with
  $\tilde{\beta}=0.01$. However, the vortex evolution were similar.}

@@

The above `planet-off' simulations are not formally linear
stability calculations, because the cooling time is comparable or shorter
than the instability growth time, $t_c\lesssim\gamma^{-1}$.  
Thus the disc cools back to its initial temperature corresponding to
$h=0.05$ before or during the instability growth, so we do not 
have a steady basic state to formulate a standard linear stability 
problem.

@@

{\bf This decay is probably due to numerical viscosity. During the
  slow decay the vortex elongates (weakens) while its radial width
  remains $O(H)$, so its surface density decreases. In addition, for 
  $\tilde{\beta}=0.1,\,0.5$ and $1.0$ we also observe the appearance of
  spiral waves associated with the vortex, which may contribute to
  its dissipation (see below).} 
% (true?) (For isothermal and long
% cases yes. Spiral waves appear in beta=0.1,0.5,1.0}) 
We will see in the
next section that this {\bf decay after linear growth} is very
different to when the planet potential is kept on.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1.1cm]{figures/longterm_planetoff}
  \caption{Long term simulations without the planet potential after
    the gap is set up. The $m=1$ surface density component,
    non-dimensionalized by the initial $m=0$ component, at the
    outer gap edge is shown as a function of the cooling timescale. 
  } \label{planetofflifetimeplot}
\end{figure}

@@

\section{Non-linear evolution of
  gap-edge vortices with finite cooling time}\label{nonlinear} 
We now examine long-term simulations of gap-edge vortices for
$\tilde{\beta}=0.1,0.5,1,5,10$.{\bf(Additional cases are presented in 
  \S\ref{lifetime_discuss} when examining vortex lifetimes as a
  function of $\tilde{\beta}$.)} The planet potential is kept on
throughout. We employ a grid with $(N_r,N_{\phi})=(512,1024)$ in order
for these 
simulations to be computationally feasible. We also use a larger
disc with $r_{\mathrm{out}}=45r_{\mathrm{in}}$ to minimise boundary
effects on vortex evolution, and apply open boundaries at
$r=r_\mathrm{in},\,r_\mathrm{out}$.

We comment that lower-resolution simulations with
$(N_r,N_{\phi})=(256,512)$ show similar behaviour and trends as the
high-resolution runs reported below. {\bf These simulations
  include a comparison between a run with $\tilde{\beta}=0.01$ with a 
  locally isothermal disc, which show similar behavior. In particular,
  we find vortex lifetimes have a non-monotonic dependence on
  $\tilde{\beta}$, with longest lifetimes as $\tilde{\beta}\to 0$ (or
  locally isothermal discs) and for $\tilde{\beta} = O(1)$}.

@@

\begin{figure}
  \includegraphics[width=\linewidth,clip=true,trim=0.5cm
  0cm 0cm 1cm]{figures/vortex_density}
  \caption{Average value of the relative surface density perturbation
    at vortex centres for various cooling
    times. Initial vortex over-densities decrease with cooling rate while
    vortex growth rates increase with cooling rate.
    \label{overdensity}}     
\end{figure}

@@

{\bf During the quasi-steady state the vortex 
  orbits at $r\sim r_p+6r_h$. We do not see significant
  vortex migration at this stage, since the vortex is located at a 
  surface density maximum \citep{paardekooper10}. However,
  simultaneous with the appearance of the wakes, we observe the vortex
  begins to migrate inwards to $r\sim r_p+5r_h$.  
  % This is a result of the the spiral waves transporting
  % angular momentum to the disc (Is this true? Shocks on the inner disc
  % seem stronger so I would think it would migrate outward). 
  }

@@

We also measured large increases in the Mach number near
the vortex as the $m=1$ surface density amplitude reaches maximum and begins to decay. 
Fig. \ref{machplot} plots the Mach number $M=|\bm{v} -
\bm{v}_\mathrm{vor}|/c_s$, where 
$\bm{v}_\mathrm{vor}$ corresponds to the bulk velocity of the vortex
around the disc. Values in Fig. \ref{machplot} have been averaged over
a region within $2H$ of the vortex centre. 
During the quasi-steady state the Mach number increases 
steadily, and for all cases $M$ maximizes about
% with larger growth rate for higher $\tilde\beta$.
$\sim 100P_0$ after the start the $m=1$ surface density
amplitude starts to decay. 
% {\bf check if  statement is true}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gapchange}
  \caption{Azimuthally averaged profiles of the relative surface
    density perturbations for  $\tilde\beta=1.0$
    before (solid) and after vortex decay
    (dashed). Alongside drastic reduction in the outer gap maxima, vortex decay
    smooth out the sharp outer gap edge.
    \label{gap_smoothed}} 
\end{figure}

We characterise the smoothness of the outer gap edge with a dimensionless
gap edge gradient parameter
\begin{align}
  \delta\Sigma(t)= \la \frac{\partial \langle \Sigma(t,r)
    \rangle_{\phi}}{\partial r} \cdot \frac{r}{\langle \Sigma(t=0,r)
    \rangle_{\phi}} \ra_{\Delta r}  
\end{align}
where $\Delta r = r\in [r_p,r_p + 6r_h]$ is the radial range of averaging, spanning
from centre of the gap to the radius of the surface density maximum.
A larger $\delta \Sigma$ characterizes a sharper gap edge and  
larger local surface density maxima.

@@

This can be interpreted as the vortex providing a viscosity;  
and we measure a typical alpha viscosity $\alpha = O(10^{-2})$
associated with the vortex. This acts against gap-opening
by the planet{\bf, and smooths out the outer surface density bump,    
} so the condition for the RWI 
becomes less favourable. {\bf In
  order to re-launch the RWI, the surface density bump should
  reform. However, this is difficult as there is no more material
  in the planet's vicinity to clear out (Fig. \ref{gap_smoothed}) to
  form a surface density bump outside the gap.} {\bf This may} explain
why vortices do  not reform again (at least within the simulation timescale). 
{\bf After full decay the aspect ratio at the outer gap edge is
  $h\sim0.05$. }
% (0.06 for beta=5.0,10)}

@@

\begin{figure}
  \includegraphics[width=\linewidth]{figures/gap_smoothness}
  \caption{Non-dimensional measure of the surface density gradient at
    the outer gap edge $\tilde\beta=1.0$. During the vortex quasi-steady state
    the gap edge is found to have a large gradient and sharp peak while vortex
    dissipation,
    which occurs at $t\approx1000P_0$ as seen in Fig. \ref{lifetimeplot},
    works to smooth out the gap edge.
% A large $\delta\Sigma$
 %   corresponds to a more sharply peaked outer gap edge.
    \label{smoothnessplot}}  
\end{figure}

@@

We now examine vortex lifetimes as function of the imposed cooling
times. {\bf For this study, additional simulations with intermediate
  values of $\tilde\beta=0.25,0.75,2.5,7.5$ were performed, as well as
  a nearly (locally) isothermal case with $\tilde{\beta}=0.01$.} 
 We define the vortex lifetime, $t_{\mathrm{life}}$, as the time
at which the over-density of the vortex
% to reach $10\%$ of its max overdensity value
returns to $\Delta \Sigma/\Sigma_0\sim1$ \emph{after} reaching maximum
%{\bf correct interpretation?}
 (which is on the order of the
initial over-density associated with the gap formation).

@@

For fast cooling rates ($\tilde{\beta}\lesssim 1$), the
vortex lifetime is maximized for $\tilde{\beta}\to0$: we find 
$t_{\mathrm{life}} \approx 2100P_0$ for $\tilde{\beta}=0.01$ and 
decreases to $t_{\mathrm{life}} \approx 1100P_0$ for
$\tilde{\beta}=0.5$. This is due to the longer decay timescale with
decreasing $\tilde{\beta}$. For longer cooling times ($\tilde{\beta}\gtrsim
1$) the vortex lifetimes maximizes at $\tilde\beta=2.5$ with 
$t_{\mathrm{life}}\approx 1650P_0$.  The other timescales $t_\mathrm{diss}$ and
$t_\mathrm{Mach}$ are less complicated, but still display a
non-monotonic dependence on $\tilde{\beta}$, with a single maximum at
$\tilde{\beta}=1$. 
% his is in comparison with the timescale of intial vortex dissipation which has 
% non-monotonic dependence and a single maxima at $\tilde\beta=5.0$.
%non-monotonic dependence for all timescales
